---
title: "Customizing Population Time Plots"
author: "Sahir R. Bhatnagar"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_height: 8
    fig_width: 11
    keep_md: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{Customizing Population Time Plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: reference.bib
editor_options: 
  chunk_output_type: console
---


# Setup

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(survival)
library(casebase)
library(ggplot2)
library(data.table)

# create poptime exposure object for ERSPC data
x <- popTime(ERSPC, time = "Follow.Up.Time", event = "DeadOfPrCa", exposure = "ScrArm")
head(x)
```



# Introduction

In this vignette, we explain in details how to customize population time plots. More specifically, we details the inner workings of the `plot` method for objects of class `popTime` and `popTimeExposure`. 


# The `.params` arguments

The user can have greater control over the aesthetics of a population plot by specifying these in the `.params` arguments. These need to be specified as lists and are subsequently passed on to the following `ggplot2` functions:

- `ribbon.params` --> [`ggplot2::geom_ribbon()`](https://ggplot2.tidyverse.org/reference/geom_ribbon.html)
- `case.params` --> [`ggplot2::geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)
- `base.params` --> [`ggplot2::geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)
- `competing.params` --> [`ggplot2::geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)
- `legend.params` --> [`ggplot2::scale_colour_manual()`](https://ggplot2.tidyverse.org/reference/scale_manual.html)
- `theme.params` --> [`ggplot2::theme()`](https://ggplot2.tidyverse.org/reference/theme.html)

Following the suggestion from the [ggplot2 book](https://ggplot2-book.org/programming.html#multiple-components), we use the `utils::modifyList` function to replace the corresponding default values. For example, the default arguments passed to the `geom_ribbon` function for plotting the area are given by

```{r, eval=TRUE}
list(data = x,
     mapping = aes(x = time, ymin = 0, ymax = ycoord),
     fill = "grey80")
```

(Note that the variables `time` and `ycoord` are columns created by the `casebase::popTime` function). Suppose we want to change the fill color. We simply specify this color in the `ribbon.params` argument:

```{r, eval=TRUE}
ribbon.params <- list(fill = "#0072B2")
```

We then call the `utils::modifyList` function to override the function defaults:

```{r, eval=TRUE}
(new_ribbon_params <- utils::modifyList(list(data = x,
                                            mapping = aes(x = time, ymin = 0, ymax = ycoord),
                                            fill = "grey80"), 
                                       ribbon.params))
```


Finally, we use `base::do.call` to execute the `geom_ribbon` function on this list:

```{r, eval=TRUE}
ggplot() + base::do.call("geom_ribbon", new_ribbon_params)
```


# Change the Facet Labels

The default arguments to the `facet.params` argument is given by:

```{r}
exposure_variable <- attr(x, "exposure")
default_facet_params <- list(facets = exposure_variable, ncol = 1)
```

The poulation time area stratified by treatment arm is then plotted using the following code

```{r}
ggplot() + 
    base::do.call("geom_ribbon", new_ribbon_params) + 
    base::do.call("facet_wrap", default_facet_params) 

# this is equivalent to
# plot(x, add.case.series = FALSE)
```

We can modify the facet labels by either changing the factor labels in the data or specifying the `labeller` argument. See [this blog post](https://sahirbhatnagar.com/blog/2016/02/08/math-expressions-with-facets-in-ggplot2/) for further details. Here is an example of how we can change the facet labels using the `plot` method provided by the casebase package:

```{r}
# Use character vectors as lookup tables:
group_status <- c(
  `0` = "Control Arm",
  `1` = "Screening Arm"
)

plot(x, 
     add.case.series = FALSE, # do not plot the case serires
     facet.params = list(labeller = labeller(ScrArm = group_status), # change labels
                         strip.position = "right") # change facet position
     ) 
```



# Changing the Plot Aesthetics

Suppose we want to change the color of the points and the legend labels. We use the `bmtcrr` dataset as the example in this section. 

The default colors for the case series, base series and competing event are given by:

```{r}
cols <- c("Case series" = "#D55E00", "Competing event" = "#009E73", "Base series" = "#0072B2")
```


## Change only the point colors

If you only want to change the color points, you must specify a named vector exactly as specified in the `cols` object created above. Note that the names `Case series`, `Base Series` and `Competing event` must remain the same, otherwise the function won't know how to map the colors to the corresponding points. This is because the `colour` aesthetic mapping in the `geom_point` functions have been set to `Case series`, `Base Series` and `Competing event`. For example, the default call to `geom_point` for the case series is given by:

```{r}
ggplot() + do.call("geom_point", list(data = x[event == 1],
                     mapping = aes(x = time, y = yc, colour = "Case series")))
```


We define a new set of colors:

```{r}
new_cols <- c("Case series" = "red", "Competing event" = "orange", "Base series" = "blue")
```

We then pass `new_cols` to the `legend.params` argument. Internally, this gets passed to the [`ggplot2::scale_color_manual`](https://ggplot2.tidyverse.org/reference/scale_manual.html) function:

```{r, eval=FALSE}
do.call("scale_colour_manual", utils::modifyList(
  list(name = element_blank(),
       breaks = c("Case series", "Competing event", "Base series"),
       values = cols), list(values = new_cols))
)
```

Here is the code to only change the colors:

```{r}
# this data ships with the casebase package
data("bmtcrr")

popTimeData <- popTime(data = bmtcrr, time = "ftime", exposure = "D")
plot(popTimeData,
     add.case.series = TRUE, 
     add.base.series = TRUE,
     add.competing.event = TRUE,
     ratio = 1,
     comprisk = TRUE,
     legend = TRUE,
     legend.params = list(values = new_cols))
```



## Change Point Color and Legend Labels

In order to change both the point colors and legend labels, we must modify the aesthetic mapping of the `geom_point` calls as follows:

```{r}
# this data ships with the casebase package
data("bmtcrr")

popTimeData <- popTime(data = bmtcrr, time = "ftime", exposure = "D")
plot(popTimeData,
     add.case.series = TRUE, 
     add.base.series = TRUE,
     add.competing.event = TRUE,
     ratio = 1,
     comprisk = TRUE,
     legend = TRUE,
     case.params = list(mapping = aes(x = time, y = yc, colour = "Relapse")),
     base.params = list(mapping = aes(x = time, y = ycoord, colour = "Base series")),
     competing.params = list(mapping = aes(x = time, y = yc, colour = "Competing event")),
     legend.params = list(name = "Legend Name",
                          breaks = c("Relapse", "Base series", "Competing event"),
                          values = c("Relapse" = "blue", "Competing event" = "hotpink", "Base series" = "orange")))
```


NOTE: the lists being passed to the `.params` arguments must be named arguments, otherwise they will give unexpected behavior. For example

```{r, eval=FALSE}
# this will work because mapping is the name of the 
# argument of the list 
case.params = list(mapping = aes(x = time, y = yc, colour = "Relapse"))
```

```{r, eval=FALSE}
# this will NOT work because the argument of the list has no name
# and therefore utils::modifyList, will not override the defaults. 
case.params = list(aes(x = time, y = yc, colour = "Relapse"))
```




# Session information

```{r echo=FALSE, eval=TRUE}
print(sessionInfo(), locale = F)
```

