---
title: "Plot Cumulative Incidence and Survival Curves"
author: "Sahir Rai Bhatnagar"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Plot Cumulative Incidence and Survival Curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: reference.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  results='hide',
  fig.show='all',
  comment = "#>"
)
```


## Introduction

In this short vignette, we will introduce the `plot` method for `absoluteRisk` objects. This method allows you to plot cumulative incidence (CI) or survival curves as a function of time and a given covariate profile. More specifically, the cumulative incidence is given by:

$$ CI(x, t) = 1 - exp\left[ - \int_0^t h(x, u) \textrm{d}u \right] $$ 
where \\( h(x, t) \\) is the hazard function, \\( t \\) denotes the numerical value (number of units) of a point in prognostic/prospective time and \\( x \\) is the realization of the vector \\( X \\) of variates based on the patient's profile and intervention (if any). And the survival function is given by 
$$ S(x, t) = 1 - CI(x,t) = exp\left[ - \int_0^t h(x, u) \textrm{d}u \right] $$ 


## Analysis of the `brcancer` dataset

To illustrate hazard function plots, we will use the breast cancer dataset which contains the observations of 686 women taken from the [`TH.data`](https://cran.r-project.org/package=TH.data) package. This dataset is also available from the `casebase` package. In the following, we will show the CI curve for several covariate profiles.


```{r eval=TRUE}
library(casebase)
library(survival)
library(ggplot2)


data("brcancer")
mod_cb_glm <- fitSmoothHazard(cens ~ estrec*log(time) + 
                                  horTh + 
                                  age + 
                                  menostat + 
                                  tsize + 
                                  tgrade + 
                                  pnodes + 
                                  progrec,
                              data = brcancer,
                              time = "time", ratio = 10)

summary(mod_cb_glm)
```



### Plotting Cumulative Incidence Curves

We can use the `plot` method for objects of class `absRiskCB`, which is returned by the `absoluteRisk` function, to plot cumulative incidence curves. For example, suppose we want to compare the cumulative incidence curves of the 1st and 50th individuals in the `brcancer` dataset:

```{r}
smooth_risk_brcancer <- absoluteRisk(object = mod_cb_glm, 
                                     newdata = brcancer[c(1,50),])

class(smooth_risk_brcancer)
plot(smooth_risk_brcancer)
```

These curves can be further customized. For example, suppose we want to change the legend title and legend keys:

```{r}
plot(smooth_risk_brcancer, 
     id.names = c("Covariate Profile 1","Covariate Profile 50"), 
     legend.title = "Type", 
     xlab = "time (days)", 
     ylab = "Cumulative Incidence (%)") 
```

The call to `plot` on a `absRiskCB` object returns a `ggplot2` object, and therefore can be used downstream with other `ggplot2` functions. For example, suppose we want to change the theme:

```{r}
plot(smooth_risk_brcancer, 
     id.names = c("Covariate Profile 1","Covariate Profile 50"), 
     legend.title = "Type", 
     xlab = "time (days)", 
     ylab = "Cumulative Incidence (%)") + theme_linedraw() 
```


### Using `graphics::matplot`

By default, the `plot` method uses `ggplot2` to produce the curves. Alternatively, you can use `graphics::matplot` by specifying `gg = FALSE`. This option is particularly useful if you want to add the cumulative incidence curve to an existing plot, e.g., adding the adjusted smooth curve to a Kaplan-Meier curve. In this example, we calculate the cumulative incidence for a typical individual in the dataset:

```{r}
cols <- c("#8E063B","#023FA5")

smooth_risk_typical <- absoluteRisk(object = mod_cb_glm, newdata = "typical")
y <- with(brcancer, Surv(time, cens))
plot(y, fun = "event", conf.int = F, col = cols[1])
plot(smooth_risk_typical, add = TRUE, col = cols[2], gg = FALSE)
legend("bottomright", 
       legend = c("Kaplan-Meier", "casebase"), 
       col = cols,
       lty = c(1, 1), 
       bg = "gray90")
```



## Survival Curves

We can also easily calculate and plot survival curves by specifying `type = 'survival'` in the call to `absoluteRisk`. The corresponding call to `plot` is the same as with cumulative incidence curves:

```{r}
smooth_surv_brcancer <- absoluteRisk(object = mod_cb_glm, 
                                     newdata = brcancer[c(1,50),],
                                     type = "survival")

plot(smooth_surv_brcancer)

smooth_surv_brcancer <- absoluteRisk(object = mod_cb_glm, 
                                     newdata = brcancer[1,],
                                     time = 90,
                                     type = "survival")

plot(smooth_surv_brcancer)

```


## Other families


```{r eval=TRUE}
mod_cb_glmnet <- fitSmoothHazard(cens ~ estrec*time + 
                                     horTh + 
                                     age + 
                                     menostat + 
                                     tsize + 
                                     tgrade + 
                                     pnodes + 
                                     progrec,
                                 data = brcancer,
                                 time = "time", 
                                 ratio = 1, 
                                 family = "glmnet")

# these all gives errors
absoluteRisk(object = mod_cb_glmnet)
absoluteRisk(object = mod_cb_glmnet, newdata = brcancer)
absoluteRisk(object = mod_cb_glmnet, newdata = brcancer[1:2,])
absoluteRisk(object = mod_cb_glmnet, newdata = brcancer[1:2,], time = c(500, 2000))
absoluteRisk(object = mod_cb_glmnet, newdata = "typical", time = c(500, 2000))
absoluteRisk(object = mod_cb_glmnet, time = c(500, 2000))

# this is what casebase is doing internally in hazard_estimation
formula_pred <- formula(delete.response(terms(mod_cb_glmnet$formula)))
newdata_matrix <- model.matrix(formula_pred, brcancer)
newdata_matrix <- newdata_matrix[,which(colnames(newdata_matrix) != "(Intercept)")]

# this is missing horThno column
head(newdata_matrix)
predict(mod_cb_glmnet, newdata_matrix, s = "lambda.min", newoffset = 0)

# inspection of glmnet coefficients, there is a coef for horThno
# which is causing the mismatch in the matrix multiplcation of x %*% beta
coef(mod_cb_glmnet)

# this works, and is actually what is used in casebase:::cv.glmnet.formula
newdata_matrix_2 <- model.matrix(update.formula(formula_pred, ~.-1), brcancer)
head(newdata_matrix_2)
head(predict(mod_cb_glmnet, newdata_matrix_2, s = "lambda.min", newoffset = 0))



table(brcancer$horTh)




X <- prepareX(brcancer)
Y <- X[,c("time","cens")]
X <- X[,-which(colnames(X) %in% c("time","cens"))]


mod_cb_glmnet <- fitSmoothHazard.fit(x = X, y = Y,
                                     time = "time", 
                                     event = "cens",
                                     formula_time = ~log(time),
                                     ratio = 10, family = "glmnet")

coef(mod_cb_glmnet, s = "lambda.min")

str(brcancer[1,])
smooth_risk_typical <- absoluteRisk(object = mod_cb_glmnet)

```



