---
title: "Fitting smooth-in-time parametric hazard functions"
author: "Maxime Turgeon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Methodological details

## First example

The first example we discuss uses the well-known ```veteran``` dataset, which is part of the ```survival``` package. As we can see below, there is almost no censoring, and therefore we can get a good visual representation of the survival function:

```{r eval=TRUE, fig.width=7, fig.height=7, fig.align='center'}
library(survival)
data(veteran)
table(veteran$status)

evtimes <- veteran$time[veteran$status == 1]
hist(evtimes, nclass=30, main='', xlab='Survival time (days)', col='gray90', probability=TRUE)
tgrid <- seq(0, 1000, by=10)
lines(tgrid, dexp(tgrid, rate=1.0/mean(evtimes)), lwd=2, lty='dashed', col='red')
```

We will first try to estimate the hazard function parametrically using some well-known regression routines. But first, we will reformat the data slightly.

```{r eval=TRUE}
veteran$prior <- factor(veteran$prior, levels = c(0, 10))
veteran$celltype <- factor(veteran$celltype, 
                           levels = c('large', 'squamous', 'smallcell', 'adeno'))
veteran$trt <- factor(veteran$trt, levels = c(1, 2))
```

Using the ```eha``` package, we can fit a Weibull form, with different values of the shape parameter. For $shape = 1$, we get an exponential distribution:

```{r eval=TRUE}
library(eha)
y <- with(veteran, Surv(time, status))

model1 <- weibreg(y ~ karno + diagtime + age + prior + celltype + trt, 
                  data = veteran, shape = 1)
summary(model1)
```

If we take $shape = 0$, the shape parameter is estimated along with the regression coefficients:

```{r eval=TRUE}
model2 <- weibreg(y ~ karno + diagtime + age + prior + celltype + trt, 
                  data = veteran, shape = 0)
summary(model2)
```

Finally, we can also fit a Cox proportional hazard:

```{r eval=TRUE}
model3 <- coxph(y ~ karno + diagtime + age + prior + celltype + trt, 
                data = veteran)
summary(model3)
```

As we can see, all three models are significant.

The method available in this package makes use of *case-base sampling*. That is, person-moments are randomly sampled across the entire follow-up time, with some moments corresponding to cases and others to controls. By sampling person-moments instead of individuals, we can then use logistic regression to fit smooth-in-time parametric hazard functions. See the previous section for more details.

First, we will look at the follow-up time by using population-time plots:

```{r eval=TRUE, fig.align='center', fig.height=7, fig.width=7}
nobs <- nrow(y)
ftime <- veteran$time
ord <- order(ftime, decreasing=TRUE)
plot(0, type='n', xlim=c(0,1000), ylim=c(0,150), xlab='Follow-up time', ylab='Population')
segments(rep(0.0, nobs), 1:nobs, ftime[ord], 1:nobs, col='gray25')
cases <- veteran$status == 1
points((ftime[ord])[cases[ord]], (1:nobs)[cases[ord]], pch=20, col='red', cex=0.5)
```



```{r eval=TRUE}
library(casebase)
library(data.table)
names(veteran)[names(veteran) == "status"] <- "event"
model4 <- fitSmoothHazard(event ~ time + karno + diagtime + age + prior +
             celltype + trt, data = veteran, ratio=100, type = "uniform")
summary(model4)

absoluteRisk(object = model4, time = 90)
mean(ftime <= 90)
```

```{r eval=TRUE}
# Fit a spline for time
library(splines)
model5 <- fitSmoothHazard(event ~ bs(time) + karno + diagtime + age + prior +
             celltype + trt, data = veteran, ratio=100, type = "uniform")
summary(model5)

absoluteRisk(object = model5, time = 90)
```

```{r eval=TRUE, fig.align='center', fig.height=5, fig.width=5}
plot(absoluteRisk(object = model4, time = 90, newdata = veteran), 
     absoluteRisk(object = model5, time = 90, newdata = veteran),
     xlab="Linear", ylab = "Splines", pch=19)
```

<!-- ## Vaccination example -->

<!-- ```{r eval=FALSE} -->
<!-- inpath <- '~/git_repositories/casebase/data/' -->
<!-- etimes <- read.csv(file.path(inpath, 'etimes.csv'), header=TRUE) -->
<!-- vtimes <- read.csv(file.path(inpath, 'vtimes.csv'), header=TRUE) -->
<!-- vtc <- vtimes[etimes[,1],2] -->
<!-- exposed <- (vtc < etimes[,2]) & (vtc >= etimes[,2] - 7.0) -->

<!-- popsize <- nrow(vtimes) -->
<!-- ncases <- nrow(etimes) -->
<!-- n.cases <- rep(1, ncases) -->
<!-- n.exposed.cases <- as.numeric(exposed) -->
<!-- n.unexposed.cases <- as.numeric(!exposed) -->

<!-- # Time-matched sampling: -->

<!-- sample.ctls = function(x) sample(1:popsize, x, FALSE) -->

<!-- ctl.case.ratio = 1 -->
<!-- n.ctls = n.cases * ctl.case.ratio -->
<!-- denom.sample = lapply(n.ctls, sample.ctls) -->
<!-- nctls = sum(n.ctls) -->
<!-- n.exposed.ctls <- rep(NA, ncases) -->

<!-- set.seed(1) -->
<!-- cc <- 0 -->
<!-- casestatus <- NULL -->
<!-- exposure <- NULL -->
<!-- matched <- NULL -->
<!-- for (i in 1:ncases) { -->
<!--     cc <- cc + 1 -->
<!--     exstatus <- (vtimes[denom.sample[[i]],2] < etimes[i,2]) & (vtimes[denom.sample[[i]],2] >= etimes[i,2] - 7.0) -->
<!--     n.exposed.ctls[i] = sum(exstatus) -->
<!--     casestatus <- c(casestatus, rep(1, n.cases[i]), rep(0, n.ctls[i])) -->
<!--     exposure <- c(exposure, rep(0, n.unexposed.cases[i]), rep(1, n.exposed.cases[i]), as.numeric(exstatus)) -->
<!--     matched <- c(matched, rep(cc, n.cases[i] + n.cases[i] * ctl.case.ratio)) -->
<!-- } -->
<!-- # cbind(casestatus, exposure, matched) -->

<!-- n.unexposed.ctls <- n.ctls - n.exposed.ctls -->

<!-- model1 <- clogit(casestatus ~ exposure + strata(as.factor(matched))) -->
<!-- summary(model1) -->

<!-- maxage <- 140/7 -->

<!-- # pdf(file.path(inpath, 'studybase.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'studybase.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="gray80",border=NA) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Pop.-time'), -->
<!--        pch=c(15), col=c('gray80'), pt.bg=c('gray80'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- idx <- order(vtimes[,2]) -->

<!-- # pdf(file.path(inpath, 'studybaseexposure.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'studybaseexposure.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="lightblue",border=NA) -->
<!-- segments(vtimes[idx,2]/7, (1:popsize), (vtimes[idx,2]+7)/7, (1:popsize), col='yellow2') -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.-time','Exposed pop.-time'), -->
<!--        pch=c(15,15), col=c('lightblue','yellow2'), pt.bg=c('lightblue','yellow2'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->

<!-- vidx <- order(vtc) -->
<!-- caseidx <- (1:popsize)[vtimes[idx,1] %in% etimes[,1]] -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'caseseries.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'caseseries.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="lightblue",border=NA) -->
<!-- segments(vtimes[idx,2]/7, (1:popsize), (vtimes[idx,2]+7)/7, (1:popsize), col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.-time','Exposed pop.-time','Case series'), -->
<!--        pch=c(15,15,20), col=c('lightblue','yellow2','red'), pt.bg=c('lightblue','yellow2','red'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'baseseries.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'baseseries.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="lightblue",border=NA) -->
<!-- segments(vtimes[idx,2]/7, (1:popsize), (vtimes[idx,2]+7)/7, (1:popsize), col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- points(etimes[matched[casestatus==0],2]/7, vtimes[unlist(denom.sample),1], pch=20, col='blue', cex=0.75) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.-time','Exposed pop.-time','Case series','Base series'), -->
<!--        pch=c(15,15,20,20), col=c('lightblue','yellow2','red','blue'), pt.bg=c('lightblue','yellow2','red','blue'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # Self-matched sampling: -->

<!-- minday <- 0 -->
<!-- maxday <- 140 -->

<!-- nbase <- 1 -->
<!-- times <- rep(NA, 2 * (ncases + nbase * ncases)) -->
<!-- casestatus <- rep(NA, 2 * (ncases + nbase * ncases)) -->
<!-- exposure <- rep(NA, 2 * (ncases + nbase * ncases)) -->
<!-- matched <- rep(NA, 2 * (ncases + nbase * ncases)) -->

<!-- set.seed(1) -->
<!-- counter <- 0 -->
<!-- casecounter <- 0 -->
<!-- for (i in 1:ncases) { -->
<!--     if (is.finite(etimes[i,2]) & etimes[i,2] > minday & etimes[i,2] <= maxday) { -->
<!--         counter <- counter + 1 -->
<!--         casecounter <- casecounter + 1 -->
<!--         times[counter] <- etimes[i,2] -->
<!--         casestatus[counter] <- 1 -->
<!--         exposure[counter] <- (etimes[i,2] - vtc[i]) > 0.0 & (etimes[i,2] - vtc[i]) < 7.0 -->
<!--         matched[counter] <- casecounter -->
<!--         # n <- rpois(1, nbase) -->
<!--         n <- nbase -->
<!--         for (j in 1:n) { -->
<!--             counter <- counter + 1 -->
<!--             u <- runif(1, 0.0, nbase) -->
<!--             times[counter] <- runif(1, min=minday, max=maxday) -->
<!--             casestatus[counter] <- 0 -->
<!--             exposure[counter] <- (times[counter] - vtc[i]) > 0.0 & (times[counter] - vtc[i]) < 7.0 -->
<!--             matched[counter] <- casecounter -->
<!--         } -->
<!--     } -->
<!--     # if (i %% 1000 == 0) -->
<!--     #     print(i) -->
<!-- } -->
<!-- # counter -->
<!-- # casecounter -->
<!-- times <- times[1:counter] -->
<!-- casestatus <- casestatus[1:counter] -->
<!-- exposure <- exposure[1:counter] -->
<!-- matched <- matched[1:counter] -->

<!-- times2 <- times^2 -->
<!-- model2 <- clogit(casestatus ~ exposure + times + times2 + strata(as.factor(matched))) -->
<!-- summary(model2) -->

<!-- model3 <- clogit(casestatus ~ exposure + bs(times) + strata(as.factor(matched))) -->
<!-- summary(model3) -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'caseseries2.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'caseseries2.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="lightblue",border=NA) -->
<!-- segments(vtimes[idx,2]/7, (1:popsize), (vtimes[idx,2]+7)/7, (1:popsize), col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.n-time','Exposed pop.-time','Case series'), -->
<!--        pch=c(15,15,20), col=c('lightblue','yellow2','red'), pt.bg=c('lightblue','yellow2','red'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'selfmatched.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'selfmatched.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="white",border=NA) -->
<!-- segments(rep(minday/7, ncases), caseidx, rep(maxday/7, ncases), caseidx, col='lightblue') -->
<!-- segments(vtc[vidx]/7, caseidx, (vtc[vidx]+7)/7, caseidx, col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.-time','Exposed pop.-time','Case series'), -->
<!--        pch=c(15,15,20), col=c('lightblue','yellow2','red'), pt.bg=c('lightblue','yellow2','red'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'baseseries2.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'baseseries2.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,popsize), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="white",border=NA) -->
<!-- segments(rep(minday/7, ncases), caseidx, rep(maxday/7, ncases), caseidx, col='lightblue') -->
<!-- segments(vtc[vidx]/7, caseidx, (vtc[vidx]+7)/7, caseidx, col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- points(times[casestatus==0]/7, caseidx[matched[casestatus==0]], pch=20, col='blue', cex=0.75) -->
<!-- legend(0.025 * maxage, 0.975 * popsize, legend=c('Unexposed pop.-time','Exposed pop.-time','Case series','Base series'), -->
<!--        pch=c(15,15,20,20), col=c('lightblue','yellow2','red','blue'), pt.bg=c('lightblue','yellow2','red','blue'), -->
<!--        cex=1, bg='white', box.col='white') -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ```{r eval=FALSE} -->
<!-- # pdf(file.path(inpath, 'baseseries3.pdf'), width=8, height=6, paper='special') -->
<!-- postscript(file.path(inpath, 'baseseries3.eps'), width=8, height=6, paper='special', horizontal=FALSE) -->
<!-- op <- par(mar = c(4.25,4.25,1,1)) -->
<!-- plot(1, 1, xlim=c(0,maxage), ylim=c(0,1000), type='n', xlab='Age (weeks)', ylab='Population') -->
<!-- rect(0,0,maxage,popsize, col="white",border=NA) -->
<!-- segments(rep(minday/7, ncases), caseidx, rep(maxday/7, ncases), caseidx, col='lightblue') -->
<!-- segments(vtc[vidx]/7, caseidx, (vtc[vidx]+7)/7, caseidx, col='yellow2') -->
<!-- points(etimes[vidx,2]/7, caseidx, pch=20, col='red', cex=0.75) -->
<!-- points(times[casestatus==0]/7, caseidx[matched[casestatus==0]], pch=20, col='blue', cex=0.75) -->
<!-- par(op) -->
<!-- dev.off() -->
<!-- ``` -->
